---
title: "2_Goseq analysis & IRE enrichment test"
author: "Yang & Steve"
date: "24/02/2020"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

---
### Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r packages}
library(limma)
library(edgeR)
library(AnnotationHub)
library(tidyverse)
library(magrittr)
library(pander)
library(ggrepel)
library(scales)
library(org.Hs.eg.db)
library(plyr)
library(ggraph)
library(tidygraph)
library(fgsea)
library(goseq)
library(org.Dr.eg.db)
library(msigdbr)
library(rWikiPathways)
theme_set(theme_bw())
panderOptions("big.mark", ",")
panderOptions("table.split.table", Inf)
panderOptions("table.style", "rmarkdown")
if (interactive()) setwd(here::here("analysis"))
```


Annotation was set up as a EnsDb object based on Ensembl Release 96.

```{r ensDb}
ah <- AnnotationHub() %>%
  subset(species == "Danio rerio") %>%
  subset(dataprovider == "Ensembl") %>%
  subset(rdataclass == "EnsDb")
ensDb <- ah[["AH69169"]]
genesGR <- genes(ensDb)
transGR <- transcripts(ensDb)
DrEns2Symbol <- genesGR %>%
    mcols() %>%
    as_tibble() %>%
    dplyr::select(gene_id, gene_name)
```

# Data load

```{r load dgeList}
dgeList <- read_rds(here::here("data","dgeList.rds"))
entrezGenes <- dgeList$genes %>%
  dplyr::filter(!is.na(entrez_gene)) %>%
  unnest(entrez_gene) %>%
  dplyr::rename(entrez_gene = entrez_gene)
topTable <- file.path(here::here("output", "topTable.csv")) %>% 
  read_csv()
topTableDE <- file.path(here::here("output", "DEgenes.csv")) %>% 
  read_csv()
```


# Databases used for testing

Hallmark and KEGG pathway gene mappings were achieved by `msigdbr`, and Wiki pathway gene mapping was downloaded by `rWikiPathways`.

```{r hallmark gene sets}
hallmark <- msigdbr("Danio rerio", category = "H")  %>% 
  left_join(entrezGenes) %>%
  dplyr::filter(!is.na(ensembl_gene_id)) %>%
  distinct(gs_name, ensembl_gene_id, .keep_all = TRUE)
hallmarkByGene <- hallmark %>%
  split(f = .$ensembl_gene_id) %>%
  lapply(extract2, "gs_name")
hallmarkByID <- hallmark %>%
  split(f = .$gs_name) %>%
  lapply(extract2, "ensembl_gene_id")
```

```{r kegg gene sets}
kegg <- msigdbr("Danio rerio", category = "C2", subcategory = "CP:KEGG")  %>% 
  left_join(entrezGenes) %>%
  dplyr::filter(!is.na(ensembl_gene_id)) %>%
  distinct(gs_name, ensembl_gene_id, .keep_all = TRUE)
keggByGene <- kegg  %>%
  split(f = .$ensembl_gene_id) %>%
  lapply(extract2, "gs_name")
keggByID <- kegg  %>%
  split(f = .$gs_name) %>%
  lapply(extract2, "ensembl_gene_id")
```

```{r wiki gene sets}
wikidownload <- downloadPathwayArchive(organism = "Danio rerio", format = "gmt") 
wiki <- gmtPathways(here::here("analysis","wikipathways-20200210-gmt-Danio_rerio.gmt"))
wikilist <- names(wiki) %>%
  lapply(function(x){
    tibble(pathway = x, entrez_gene = wiki[[x]])
  }) %>%
  bind_rows() %>%
  mutate(entrez_gene = as.numeric(entrez_gene)) %>%
  left_join(entrezGenes) %>%
  dplyr::filter(!is.na(ensembl_gene_id)) %>%
  distinct(pathway, ensembl_gene_id, .keep_all = TRUE) %>%
  mutate(pathway = str_remove_all(pathway, "%.+"))
wikiByGene <- wikilist  %>%
  split(f = .$ensembl_gene_id) %>%
  lapply(extract2, "pathway")
wikiByID <- wikilist  %>%
  split(f = .$pathway) %>%
  lapply(extract2, "ensembl_gene_id")
```

# GOseq analysis

Calculate pwf using gene lengths as bia data.

```{r pwf}
pwf <- topTable %>% 
  dplyr::select(ensembl_gene_id, logFC, DE, aveLength, aveGc) %>%
  mutate(nGC = aveLength*aveGc) %>% 
  distinct(ensembl_gene_id, .keep_all = TRUE) %>% 
  with(
    nullp(
      DEgenes = structure(
        as.integer(DE), names = ensembl_gene_id
      ), 
      genome = "danRer10", 
      id = "ensGene", 
      # bias.data = log(nGC),
      bias.data = aveLength,
      plot.fit = FALSE
    )
  )
```

```{r, echo=FALSE, fig.cap="Probability weight function for a gene being considered as DE based on the number of GC bases per gene."}
grid::grid.newpage()
plotPWF(pwf)
```


## Goseq pathway analysis

Use `goseq` to perform pathway enrichment analysis and print out the top 5 significantly-changed pathways

### Hallmark Gene Sets


```{r hallmark analysis}
hallmarkGoseq <- pwf %>%
    goseq(gene2cat = hallmarkByGene) %>%
    as_tibble %>%
    mutate(FDR = p.adjust(over_represented_pvalue, method = "fdr"))

hallmarkGoseq %>%
    dplyr::slice(1:5) %>%
    dplyr::select(
        Pathway = category, 
        starts_with("num"), 
        PValue = over_represented_pvalue, 
        FDR
    ) %>%
    pander(
        caption = "Most highly ranked hallmark pathways.",
        justify = "lrrrr"
    )
```

### KEGG Pathways

```{r kegg analysis}
keggGoseq <- pwf %>%
  goseq(gene2cat = keggByGene) %>%
  as_tibble %>%
  mutate(FDR = p.adjust(over_represented_pvalue, method = "fdr"))

keggGoseq %>%
    dplyr::slice(1:5) %>%
    dplyr::select(
        Pathway = category, 
        starts_with("num"), 
        PValue = over_represented_pvalue, 
        FDR
    ) %>%
    pander(
        caption = "Most highly ranked KEGG pathways.",
        justify = "lrrrr"
    )
```

### wiki pathway

```{r wiki analysis}
wikiGoseq <- pwf %>%
  goseq(gene2cat = wikiByGene) %>%
  as_tibble %>%
  mutate(FDR = p.adjust(over_represented_pvalue, method = "fdr"))

wikiGoseq %>%
    dplyr::slice(1:5) %>%
    dplyr::select(
        Pathway = category, 
        starts_with("num"), 
        PValue = over_represented_pvalue, 
        FDR
    ) %>%
    pander(
        caption = "Most highly ranked wiki pathways.",
        justify = "lrrrr"
    )
```

## Goseq GO analysis

```{r GO analysis}
goByGene <- links(org.Dr.egGO2ALLEGS) %>%
  as_tibble() %>%
  left_join(
    links(org.Dr.egENSEMBL2EG)
  ) %>%
  distinct(ensembl_id, go_id) %>%
  dplyr::filter(ensembl_id %in% topTable$ensembl_gene_id) %>%
  split(f = .$ensembl_id) %>%
  lapply(extract2, "go_id")

goGoseq <- pwf %>%
  goseq(gene2cat = goByGene) %>%
  as_tibble %>%
  mutate(FDR = p.adjust(over_represented_pvalue, method = "fdr"))
```

Enriched GO list is filter by a FDR cutoff of 0.05.

```{r enriched GO list}
goGoseq %>%
    dplyr::filter(FDR < 0.05) %>%
    dplyr::select(
        Pathway = category, 
        starts_with("num"), 
        term,
        ontology,
        PValue = over_represented_pvalue, 
        FDR
    ) %>%
    pander(
        caption = "Most highly ranked GO pathways.",
        justify = "lrrrrrr"
    )
goGoseqTop <- goGoseq %>%
    dplyr::filter(FDR < 0.05) %>%
    dplyr::select(
        Pathway = category, 
        starts_with("num"), 
        term,
        ontology,
        PValue = over_represented_pvalue, 
        FDR)
```

## GO plot

Use the links between DE genes and signficant GO terms to plot the GO networks.

```{r GO plot}
## Get significant GO terms
sigGo <- goGoseq %>%
    dplyr::filter(FDR < 0.05) %>%
    .$category

## Convert list of GO terms by gene to list of genes by GO term
geneByGo <- names(goByGene) %>% 
    lapply(function(x){tibble(gene_id = x, go_id = goByGene[[x]])}) %>% 
    bind_rows() %>% 
    split(f = .$go_id) %>% 
    lapply(magrittr::extract2, "gene_id")

## Get DE genes that belong to sigificant GO terms
goGenes <- lapply(
    sigGo, 
    function(x){
        geneByGo[[x]][geneByGo[[x]] %in% topTableDE$ensembl_gene_id]
    }
) 
names(goGenes) <- sigGo

## Make tibble of GO terms
goTerms <- names(goGenes) %>%
    tibble::enframe(name = NULL, value = "label")
## Make tibble of genes
genes <- unlist(goGenes) %>% 
    unique() %>%
    tibble::enframe(name = NULL, value = "label") %>%
    mutate

## Join to create node list
nodes <- rbind(goTerms, genes) %>%
    rowid_to_column("id")

## Create edge list
edges <- goGenes %>%
    stack() %>%
    as_tibble() %>%
    dplyr::select(goTerm = ind, geneId = values) %>%
    dplyr::arrange(goTerm) %>%
    mutate(goTerm = as.character(goTerm)) %>%
    left_join(nodes, by = c("goTerm" = "label")) %>%
    dplyr::rename(from = id) %>%
    left_join(nodes, by = c("geneId" = "label")) %>%
    dplyr::rename(to = id) %>%
    dplyr::select(from, to)

## Setup colours
colours <- length(sigGo) %>%
    rainbow()

## Create tidygraph object
tidy <- tbl_graph(
    nodes = nodes, 
    edges = edges, 
    directed = FALSE
) %>%
    activate(nodes) %>%
    mutate(
        goTerms = case_when(
            id <= length(sigGo) ~ label
        ),
        term = Term(label),
        gene_id = case_when(
            !label %in% sigGo ~ label
        ),
        colour = case_when(
            id <= length(sigGo) ~ colours[id]
        ),
        size = ifelse(id <= length(sigGo), 4, 1)
    ) %>%
    left_join(DrEns2Symbol) %>%
    activate(edges) %>%
    mutate(
        colour = case_when(
            from <= length(sigGo) ~ colours[from]
        )
    )

## Set seed to allow same graph to be produced each time function is executed
set.seed(1234)

## Plot network graph
ggraph(tidy, layout = "fr") +
    scale_fill_manual(
        values = "white", 
        na.value = "gray80"
    ) +
    scale_edge_color_manual(
        values = "black", 
        na.value = "gray80"
    ) +
    geom_edge_arc(
        aes(color = "black"), 
        alpha = 0.5, 
        show.legend = FALSE, 
        curvature = 0.5
    ) +
    geom_node_point(
        aes(fill = "black", size = size),
        shape = 21,
        stroke = 0.5, 
        show.legend = FALSE
    ) +
    geom_node_label(
        aes(label = gene_name),
        repel = TRUE,
        size = 3,
        alpha = 0.7,
        label.padding = 0.1
    ) +
    theme_graph() +
    theme(legend.position = "none")
```

# IRE enrichment analysis

## IRE Goseq analysis

```{r load IRE information}
ireGR <- list.files(
  path = here::here("analysis"),
  pattern = "gff.gz",
  full.names = TRUE
  ) %>%
    sapply(rtracklayer::import.gff, simplify = FALSE) %>%
    lapply(function(x){
        tibble(
            tx_id = seqnames(x) %>% as.character(),
            quality = x$quality
        ) %>%
            left_join(
                mcols(transGR) %>% as.data.frame()
            ) %>%
            mutate(
                quality = factor(
                    quality, levels = c("Low", "Medium", "High")
                )
            ) %>%
            arrange(gene_id, desc(quality)) %>%
            distinct(gene_id, .keep_all = TRUE) %>%
            dplyr::select(tx_id, gene_id, quality)
    }) 
names(ireGR) <- str_extract(names(ireGR), "utr[35]")

ireHigh <- lapply(ireGR, subset, quality == "High")
```


```{r IREgoseq analysis}
ireByGene <- c(
    names(ireGR) %>%
        lapply(function(x){
            mutate(ireGR[[x]], Type = paste0(x, "_All"))
        }),
    names(ireHigh) %>%
        lapply(function(x){
            mutate(ireHigh[[x]], Type = paste0(x, "_High"))
        })
) %>%
    bind_rows() %>%
    split(f = .$gene_id) %>%
    lapply(function(x){
        unique(x$Type)
    }) 

ireGoseq <- pwf %>%
   goseq(gene2cat = ireByGene) %>%
   as_tibble %>%
   mutate(FDR = p.adjust(over_represented_pvalue, method = "fdr"))
```

```{r enrichmentlist, echo=FALSE}
ireGoseq %>%
    dplyr::slice(1:5) %>%
    dplyr::select(
        Pathway = category, 
        starts_with("num"), 
        PValue = over_represented_pvalue, 
        FDR
    ) %>%
    pander(
        caption = "GOseq analysis for IRE enrichment",
        justify = "lrrrr"
    )
```

## IRE GSEA analysis 

### Gene ranks

```{r rank genes}
ranks <- topTable %>%
  mutate(stat = -sign(logFC) * log10(PValue)) %>%
  dplyr::arrange(stat) %>%
  with(structure(stat, names = ensembl_gene_id))
```

```{r IRE GSEA}
ireGSEA <- c(
    names(ireGR) %>%
        lapply(function(x){
            mutate(ireGR[[x]], Type = paste0(x, "_All"))
        }),
    names(ireHigh) %>%
        lapply(function(x){
            mutate(ireHigh[[x]], Type = paste0(x, "_High"))
        })
) %>%
    bind_rows() %>%
    split(f = .$Type) %>%
    lapply(function(x){
        unique(x$gene_id)
    }) 

fgseaIRE <- fgsea(ireGSEA, ranks, nperm=1e5) %>%
  as_tibble() %>%
  dplyr::rename(FDR = padj) %>%
  mutate(padj = p.adjust(pval, "bonferroni")) %>%
  dplyr::arrange(pval)

fgseaIRE %>%
  dplyr::select(-leadingEdge, -nMoreExtreme) %>%
  pander(
    style = "rmarkdown", 
    split.tables = Inf, 
    justify = "lrrrrrr", 
    caption = paste(
      "The", nrow(.), "GSEA analysis of IRE enrichment", percent(max(.$FDR)))
  )
```

# Data export
```{r data export}
Goseqpathway <- bind_rows(
  hallmarkGoseq,
  keggGoseq,
  wikiGoseq
) %>%
  dplyr::filter(FDR < 0.05) %>%
  dplyr::select(
    category, numDEInCat, numInCat, FDR
  ) 
write_csv(Goseqpathway, here::here("output","Goseq_pathway.csv"))

# output IRE enrichment results

write_csv(ireGoseq,here::here("output","ireGoseq_resulst.csv"))

fgseaIREresult <- fgseaIRE %>%
  dplyr::select(
    pathway, ES, NES, size, padj
  ) 
write_csv(fgseaIREresult,here::here("output","ireGSEA_resulst.csv"))
```
