---
title: "3_GSEA"
author: "Yang & Steve"
date: "24/02/2020"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---
### Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r packages}
library(limma)
library(edgeR)
library(tidyverse)
library(magrittr)
library(pander)
library(ggrepel)
library(scales)
library(plyr)
library(ggraph)
library(tidygraph)
library(fgsea)
library(pathview)
library(msigdbr)
library(rWikiPathways)
theme_set(theme_bw())
panderOptions("big.mark", ",")
panderOptions("table.split.table", Inf)
panderOptions("table.style", "rmarkdown")
if (interactive()) setwd(here::here("analysis"))
```

# Data load

```{r loaddgeList}
dgeList <- read_rds(here::here("data","dgeList.rds"))
entrezGenes <- dgeList$genes %>%
  dplyr::filter(!is.na(entrez_gene)) %>%
  unnest(entrez_gene) %>%
  dplyr::rename(entrez_gene = entrez_gene)
topTable <- file.path(here::here("output", "topTable.csv")) %>% 
  read_csv()
```

# Gene ranks

Genes were ranked by -sign(logFC)*log10(PValue).

```{r rank genes}
ranks <- topTable %>%
  mutate(stat = -sign(logFC) * log10(PValue)) %>%
  dplyr::arrange(stat) %>%
  with(structure(stat, names = ensembl_gene_id))
```


# Databases used for testing

Hallmark and KEGG pathway gene mappings were achieved by `msigdbr`, and Wiki pathway gene mapping was downloaded by `rWikiPathways`.

```{r hallmark gene set}
hallmark <- msigdbr("Danio rerio", category = "H")  %>% 
  left_join(entrezGenes) %>%
  dplyr::filter(!is.na(ensembl_gene_id)) %>%
  distinct(gs_name, ensembl_gene_id, .keep_all = TRUE)
hallmarkByGene <- hallmark %>%
  split(f = .$ensembl_gene_id) %>%
  lapply(extract2, "gs_name")
hallmarkByID <- hallmark %>%
  split(f = .$gs_name) %>%
  lapply(extract2, "ensembl_gene_id")
```

```{r kegg gene sets}
kegg <- msigdbr("Danio rerio", category = "C2", subcategory = "CP:KEGG")  %>% 
  left_join(entrezGenes) %>%
  dplyr::filter(!is.na(ensembl_gene_id)) %>%
  distinct(gs_name, ensembl_gene_id, .keep_all = TRUE)
keggByGene <- kegg  %>%
  split(f = .$ensembl_gene_id) %>%
  lapply(extract2, "gs_name")
keggByID <- kegg  %>%
  split(f = .$gs_name) %>%
  lapply(extract2, "ensembl_gene_id")
```

```{r wiki gene sets}
wikidownload <- downloadPathwayArchive(organism = "Danio rerio", format = "gmt") 
wiki <- gmtPathways(here::here("analysis", "wikipathways-20200210-gmt-Danio_rerio.gmt"))
wikilist <- names(wiki) %>%
  lapply(function(x){
    tibble(pathway = x, entrez_gene = wiki[[x]])
  }) %>%
  bind_rows() %>%
  mutate(entrez_gene = as.numeric(entrez_gene)) %>%
  left_join(entrezGenes) %>%
  dplyr::filter(!is.na(ensembl_gene_id)) %>%
  distinct(pathway, ensembl_gene_id, .keep_all = TRUE) %>%
  mutate(pathway = str_remove_all(pathway, "%.+"))
wikiByGene <- wikilist  %>%
  split(f = .$ensembl_gene_id) %>%
  lapply(extract2, "pathway")
wikiByID <- wikilist  %>%
  split(f = .$pathway) %>%
  lapply(extract2, "ensembl_gene_id")
```

# Gene Set Enrichment analysis (GSEA)

Enrichment analysis of each pathway data sets were performed by `fgsea`, using gene ranks and mappings. A bonferroni-correction cutoff of 0.05 was used to identify signficant results.

## Hallmark pathways

```{r hallmark analysis}
set.seed(22)
# Run GSEA for hallmark
fgseaHallmark <- fgsea(hallmarkByID, ranks, nperm=1e5) %>%
  as_tibble() %>%
  dplyr::rename(FDR = padj) %>%
  mutate(padj = p.adjust(pval, "bonferroni")) %>%
  dplyr::arrange(pval)

fgseaHallmarkTop <- fgseaHallmark %>%
  dplyr::filter(padj < 0.05) 

fgseaHallmarkTop %>%
  dplyr::select(-leadingEdge, -nMoreExtreme) %>%
  pander(
    style = "rmarkdown", 
    split.tables = Inf, 
    justify = "lrrrrrr", 
    caption = paste(
      "The", nrow(.), "most significantly enriched Hallmark pathways.",
      "This corresponds to an FDR of", percent(max(.$FDR)))
  )
```


```{r fig.align = "center", fig.width=12}
# Make a table plot of significant Hallmark pathways
if (interactive()) grid::grid.newpage()
plotGseaTable(
  hallmarkByID[dplyr::filter(fgseaHallmark, padj < 0.05)$pathway], ranks, fgseaHallmark, gseaParam = 0.5
)
```

## KEGG pathways

```{r KEGG analysis}
# Set seed to enable reproducibility
set.seed(22)
# Run GSEA for KEGG
fgseaKEGG <- fgsea(keggByID, ranks, nperm=1e5) %>%
  as_tibble() %>%
  dplyr::rename(FDR = padj) %>%
  mutate(padj = p.adjust(pval, "bonferroni")) %>%
  dplyr::arrange(pval)
# Create an object of pathways with adjusted p-value < 0.05 for construction of network diagrams. This should be done differently next time, but too much work has been done to change it now.
fgseaKEGGTop <- fgseaKEGG %>%
  dplyr::filter(padj < 0.05)
fgseaKEGGTop %>%
  dplyr::select(-leadingEdge, -nMoreExtreme) %>%
  pander(
    style = "rmarkdown", 
    split.tables = Inf, 
    justify = "lrrrrrr", 
    caption = paste(
      "The", nrow(.), "most significantly enriched KEGG pathways.",
      "This corresponds to an FDR of", percent(max(.$FDR)))
  )
```

```{r fig.align = "center", fig.width=10}
# Make a table plot of significant KEGG pathways
if (interactive()) grid::grid.newpage()
plotGseaTable(
  keggByID[fgseaKEGGTop$pathway], ranks, fgseaKEGG, gseaParam = 0.5
)
```

## WikiPathways

```{r WikiPathways analysis}
# Set seed to enable reproducibility
set.seed(22)
# Run GSEA for WikiPathways
fgseaWiki <- fgsea(wikiByID, ranks, nperm=1e5) %>%
  as_tibble() %>%
  dplyr::rename(FDR = padj) %>%
  mutate(padj = p.adjust(pval, "bonferroni")) %>%
  dplyr::arrange(pval)
# Create an object of pathways with adjusted p-value < 0.05 for construction of network diagrams. This should be done differently next time, but too much work has been done to change it now.
fgseaWikiTop <- fgseaWiki %>%
  dplyr::filter(padj < 0.05)
fgseaWikiTop %>%
  dplyr::select(-leadingEdge, -nMoreExtreme) %>%
  pander(
    style = "rmarkdown", 
    split.tables = Inf, 
    justify = "lrrrrrr", 
    caption = paste(
      "The", nrow(.), "most significantly enriched Wiki pathways.",
      "This corresponds to an FDR of", percent(max(.$FDR)))
  )
```

```{r fig.align = "center"}
# Make a table plot of significant WikiPathways pathways
if (interactive()) grid::grid.newpage()
plotGseaTable(
  wikiByID[fgseaWikiTop$pathway], ranks, fgseaWiki, gseaParam = 0.5
)
```

# Data export

```{r }
GSEAresult <- bind_rows(
  fgseaHallmark,
  fgseaKEGG,
  fgseaWiki
) %>%
  dplyr::filter(padj < 0.05) %>%
  dplyr::select(
    pathway, ES, NES, size, padj
  ) 
write_csv(GSEAresult,here::here("output","GSEA_resulst.csv"))
```
Use `pathview` to plot each significant KEGG pathway so as to visualize the changes.

```{r plot_keggdiagram}
keggDir <- here::here("keggdiagram")
keggPaths <- list.files(keggDir, pattern = "xml") %>% 
  str_replace_all(pattern = "dre([0-9]+).xml", "\\1")
```


```{r pv.out, results='hide'}
pv.out <- pathview(
  gene.data = topTable %>%
    dplyr::select(c("ensembl_gene_id", "logFC")) %>%
    as.data.frame() %>%
    column_to_rownames("ensembl_gene_id"),
  pathway.id = keggPaths,
  species = "Danio rerio",
  kegg.dir = keggDir,
  gene.idtype = "ENSEMBL",
  limit = list(gene = 0.8, cpd = 0.8),
  bins = list(gene = 16, cpd = 16)
)
```


```{r movePngFiles, results='hide'}
keggPng <- list.files(pattern = "dre.+pathview.png", full.names = TRUE)
file.rename(
  keggPng,
  here::here("docs", "figure", "3_GSEA.Rmd", basename(keggPng))
)
```

All generated KEGG pathview files are available using the following links:

```{r pngUrl}
pngUrl <- tibble(
  path = here::here("docs", "figure", "3_GSEA.Rmd", basename(keggPng)),
  id = str_extract(path, "dre[0-9]+")
) %>%
  split(f = .$id) %>%
  vapply(function(x){
    paste0("![](", x$path, ")\n\n\n")
  },
  character(1))
```


```{r, echo=FALSE, results='asis'}
cat(pngUrl)
```



