---
title: "Home"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(
  autodep = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

```{r loadPackages}
library(tidyverse)
```

This website contains the analysis of RNAseq data from an experiment comparing 7 dpf zebrafish larvae from two genotypes.
One genotype represented wild-type larvae, whilst the other genotype was a heterozygous mutation for the gene *psen1*, in which 6 nucleotides were deleted, to remove the two amino acids Q96 and K97.
This mutation is generally referred to as *psen1^Q96K97^*.

Results from all steps can be found at:

1. [DE Gene Analysis](1_DE_gene_analysis.html)
2. [Goseq Analysis & IRE enrichment test](2_Goseq_analysis_and_IRE_enrichment_test.html)
3. [GSEA](3_GSEA.html)
4. [DE gene omparison with 6month brain and validating qPCR plot](4_DE_gene_comparison_with_6month_brain_and_validating_qPCR_plot.html)
5. [WGCNA](5_WGCNA.html)
6. [Variance partition analysis](6_Variance_Partition_Analysis.html)
